Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-10-24 paper/main.tex
added some explanatory text for the predictions analyses; also updated y-axes of predict-knowledge-questions fig

@@

Finally, after watching \textit{Birth of Stars}, estimated knowledge for
held-out correctly answered questions (from the third quiz;
Fig.~\ref{fig:predictions}, right column) was higher for held-out correctly
answered questions than for held-out incorrectly answered questions. This held
when we included all questions in the analysis ($\U = 38279, p = 0.022$), when
we carried out across-lecture predictions (\textit{Four Fundamental Forces}: $\U
= 6684.5, p = 0.032$; \textit{Birth of Stars}: $\U = 6414.5, p = 0.002$), and
and when we carried out within-lecture predictions of held-out \textit{Birth of
Stars} questions using other \textit{Birth of Stars} questions from the same
quiz and participant ($\U = 6126, p = 0.006$). However, we found the
\textit{opposite} effect when we carried out within-lecture predictions of
held-out \textit{Four Fundamental Forces} questions using other \textit{Four
Fundamental Forces} questions from the same quiz and participant ($\U = 6734, p
= 0.027$). Specifically, held-out correctly answered Quiz 3 questions about
\textit{Four Fundamental Forces} had reliably \textit{lower} estimated
knowledge than held-out incorrectly answered questions. Speculatively, we
suggest that this may reflect participants forgetting some of the \textit{Four
Fundamental Forces} content. If this forgetting happens in a relatively
``random'' way (with respect to spatial distance within the text embedding
space), then it could explain why some held-out questions about \textit{Four
Fundamental Forces} were answered incorrectly, even if questions at nearby
coordinates (i.e., about similar content) were answered correctly. This might
lead our approach to over-estimate knowledge for held-out questions about
``forgotten'' knowledge that participants answered incorrectly.

>>>>

We carried out these analyses in three different ways. First, we used all of
the questions from a given quiz (and participant) to predict knowledge at the
embedding coordinate of a held-out question (``All questions'' in
Fig.~\ref{fig:predictions}). This test was intended to serve as an overall
baseline for the predictive power of our approach. Second, we used questions
from one lecture to predict knowledge at the embedding coordinate of a held-out
question about the \textit{other} lecture from the same quiz and participant
(``Across-lecture'' in Fig.~\ref{fig:predictions}). This test was intended to
test the \textit{generalizability} of our approach by asking whether our
knowledge predictions held across the content areas of the two lectures. Third,
we used questions from one lecture to predict knowledge at the embedding
coordinate of a held-out question about the \textit{same} lecture from the same
quiz and participant (``Within-lecture'' in Fig.~\ref{fig:predictions}). This
test was intended to test the \textit{specificity} of our approach by asking
whether our knowledge predictions could distinguish between questions about
different content covered by the same lecture. We repeated each of these
analyses using all possible held-out questions for each quiz and participant.

@@

Finally, after watching \textit{Birth of Stars}, estimated knowledge for
held-out correctly answered questions (from the third quiz;
Fig.~\ref{fig:predictions}, right column) was higher for held-out correctly
answered questions than for held-out incorrectly answered questions. This held
when we included all questions in the analysis ($\U = 38279, p = 0.022$), when
we carried out across-lecture predictions (\textit{Four Fundamental Forces}: $\U
= 6684.5, p = 0.032$; \textit{Birth of Stars}: $\U = 6414.5, p = 0.002$), and
and when we carried out within-lecture predictions of held-out \textit{Birth of
Stars} questions using other \textit{Birth of Stars} questions from the same
quiz and participant ($\U = 6126, p = 0.006$). However, we found the
\textit{opposite} effect when we carried out within-lecture predictions of
held-out \textit{Four Fundamental Forces} questions using other \textit{Four
Fundamental Forces} questions from the same quiz and participant ($\U = 6734, p
= 0.027$). Specifically, held-out correctly answered Quiz 3 questions about
\textit{Four Fundamental Forces} had reliably \textit{lower} estimated
knowledge than held-out incorrectly answered questions. Speculatively, we
suggest that this may reflect participants forgetting some of the \textit{Four
Fundamental Forces} content. If this forgetting happens in a relatively
``random'' way (with respect to spatial distance within the text embedding
space), then it could explain why some held-out questions about \textit{Four
Fundamental Forces} were answered incorrectly, even if questions at nearby
coordinates (i.e., about similar content) were answered correctly. This might
lead our approach to over-estimate knowledge for held-out questions about
``forgotten'' knowledge that participants answered incorrectly.  Taken together,
the results in Figure~\ref{fig:predictions} indicate that our approach can
reliably predict acquired knowledge (especially about recently learned content),
and that the knowledge predictions are generalizable across the content areas
spanned by the two lectures and also specific enough to distinguish between
questions about different content within the same lecture.
